---
title: "Local Burden and Demographic Disparities in Alzheimer's Disease and Related Dementias: A Generalizability Analysis of Stanford ADRC Clinical Core Data"
author:
  - name: Tracy Lam-Hine, DrPH, MBA
    affil: 1
  - name: Lisa Goldman-Rosas, PhD, MPH
    affil: "1,2"
  - name: Victor W. Henderson, MD, MS
    affil: "1,3"
affiliation:
  - num: 1
    address: Department of Epidemiology and Population Health, Stanford School of Medicine
  - num: 2
    address: Department of Primary Care and Population Health, Stanford School of Medicine
  - num: 3
    address: Department of Neurology and Neurological Sciences, Stanford School of Medicine
poster_height: "36in"
poster_width: "54in"
primary_colour: "#8C1515"
secondary_colour: "#4D4F53"
accent_colour: "#B1040E"
title_textsize: "60pt"
author_textsize: "36pt"
affiliation_textsize: "24pt"
body_textsize: "28pt"
caption_textsize: "20pt"
titlebox_height: "9%"
column_numbers: 3
output:
  posterdown::posterdown_html:
    self_contained: true
    pandoc_args: --mathjax
bibliography: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.align = "center", out.width = "100%"
)
library(tidyverse)
library(knitr)
library(kableExtra)
results_dir <- "/Users/lamhine/Documents/GitHub/gen_sadrc_clin/03_results"
hybrid     <- readRDS(file.path(results_dir, "12_hybrid_prevalence.rds"))
pr_pd      <- readRDS(file.path(results_dir, "09_pr_pd_results.rds"))
aipsw      <- readRDS(file.path(results_dir, "07_aipsw_results.rds"))
race_colors <- c("White" = "#8C1515", "Hispanic" = "#E98300",
                  "Asian" = "#007C92", "Other" = "#53565A")
race_relabel <- c("White NH" = "White", "Asian NH" = "Asian", "NH Other" = "Other", "Hispanic" = "Hispanic")
```

```{css, echo=FALSE}
@page { size: 54in 36in; margin: 0; }
html, body { overflow: hidden; max-height: 36in; }
h1, h2, h3 { break-after: avoid; }
.poster_title {
  padding-bottom: 0.2em !important;
  padding-top: 0.2em !important;
  min-height: 0 !important;
  height: auto !important;
}
.poster_title h1 { margin-bottom: 0.1em !important; margin-top: 0 !important; }
.poster_title h3, .poster_title h5 { margin-top: 0.1em !important; margin-bottom: 0.1em !important; }
.poster_body { padding-top: 0.3em !important; }
.poster_title h3, .poster_title .author, .poster_title sup { color: white !important; }
.poster_body p { margin-bottom: 0.3em; }
.poster_body h1 { margin-bottom: 0.2em; }
.poster_body h3 { margin-bottom: 0.2em; margin-top: 0.3em; }
.poster_body li { margin-bottom: 0.5em; }
```

# Background

- **Alzheimer's disease (AD) and AD-related dementias (AD/ADRD)** affect over 6 million US adults. National studies estimate ~9% AD prevalence (Dhana et al. 2023) and ~10% ADRD prevalence (Matthews et al. 2019) among adults 65+, but local estimates remain limited.

- **National models often lack estimates for Asian adults** -- a critical gap for the Bay Area, where Asian residents represent a substantial share of the older adult population.

- **Local estimates** are needed for resource allocation and caregiver support planning, but most existing estimates come from national models that may not reflect regional demographics.

- **Alzheimer's Disease Research Centers (ADRCs)** provide rich clinical data but over-enroll people with cognitive impairment. This **clinical enrichment** complicates generalization of ADRC findings.

```{r dag, fig.height=2.2, fig.width=5, fig.align='center', out.width="65%"}
dag_nodes <- tibble(
  label = c("X\n(Demographics)", "Y\n(Cognitive Status)", "S = 1\n(ADRC Enrollment)"),
  x = c(0, 2, 1), y = c(1, 1, 0))
dag_edges <- tibble(
  x = c(0, 0, 2), y = c(1, 1, 1), xend = c(2, 1, 1), yend = c(1, 0, 0),
  label = c("Causal", "Selection", "Selection\n(enrichment)"))
ggplot() +
  geom_segment(data = dag_edges, aes(x=x, y=y, xend=xend, yend=yend),
               arrow = arrow(length=unit(0.12,"inches"), type="closed"),
               linewidth=0.7, color="#4D4F53") +
  geom_text(data = dag_edges,
            aes(x=(x+xend)/2 + c(0,-0.22,0.22), y=(y+yend)/2 + c(0.1,0,0), label=label),
            size=3, color="#4D4F53", fontface="italic") +
  geom_point(data = dag_nodes, aes(x=x, y=y), size=16, shape=21,
             fill="white", color="#8C1515", stroke=1.5) +
  geom_text(data = dag_nodes, aes(x=x, y=y, label=label),
            size=2.8, fontface="bold", lineheight=0.85) +
  annotate("rect", xmin=0.55, xmax=1.45, ymin=-0.28, ymax=0.28,
           fill=NA, color="#B1040E", linewidth=1, linetype="dashed") +
  annotate("text", x=1, y=-0.38, label="Conditioned on (observed sample)",
           size=2.5, color="#B1040E", fontface="italic") +
  coord_fixed(ratio=0.7, xlim=c(-0.5,2.5), ylim=c(-0.5,1.2)) +
  theme_void()
```

### Research Questions

1. What is the estimated local prevalence of AD/ADRD among adults aged 65+ in the ADRC's three-county catchment area?
2. Are there disparities in cognitive outcomes across racial/ethnic groups after demographic standardization?
3. How does clinical enrichment shape the ADRC sample relative to the target population?

# Data & Methods

- **Data:** We used data from the Stanford ADRC (529 adults 65+, NACC UDS v3, 2015--2024), California BRFSS (2,150 adults 65+ in Alameda, San Mateo, and Santa Clara counties, 2015--2024), and published coefficients from Dhana et al. (2023) and Matthews et al. (2019).

- **Q1 -- Local AD/ADRD Prevalence:** Dhana et al. published logistic regression coefficients for AD prevalence, but their model lacks an Asian category. We constructed a hybrid model:

    $$\text{logit}(P(\text{AD}_i)) = \hat{\alpha}^{D} + \hat{\beta}_{\text{age}}^{D} x_{\text{age}} + \hat{\beta}_{\text{sex}}^{D} x_{\text{sex}} + \hat{\beta}_{\text{edu}}^{D} x_{\text{edu}} + \hat{\beta}_{\text{race}}^{A} x_{\text{race}}$$

    where $D$ denotes published CHAP coefficients and $\hat{\beta}_{\text{race}}^{A}$ are log-odds ratios for AD by race/ethnicity (vs. White) from the Stanford ADRC. We verified transportability of these ORs by testing for variation in clinical enrichment across groups. We applied the model to individual-level BRFSS data with survey weights. For ADRD, we used Matthews et al. (2019) rates via indirect standardization.

- **Q2 -- Disparities (AIPSW):** We analyzed AD, ADRD, and mild cognitive impairment (MCI) as separate outcomes. We estimated prevalence ratios using augmented inverse probability of selection weighting (Dahabreh et al. 2019), a doubly robust estimator combining outcome models with stabilized inverse odds of selection weights. We estimated variance via stratified bootstrap (1,000 reps $\times$ 40 imputations, Rubin's rules).

- **Q3 -- Enrichment:** We characterized ADRC referral pathways and compared ADRC-derived odds ratios to published population-based estimates to verify that enrichment is approximately constant across groups -- a key assumption for transportability of the disparity analysis.

# Results

```{r prevalence-plot, fig.height=4, fig.width=8}
prev_race <- hybrid$by_race %>%
  select(race_eth4_h, hybrid_prev) %>%
  rename(race = race_eth4_h, prevalence = hybrid_prev) %>%
  mutate(type = "AD (hybrid model)",
         race = recode(race, !!!race_relabel),
         race = factor(race, levels = c("White","Hispanic","Asian","Other")))
adrd_race <- hybrid$adrd_by_race %>%
  rename(race = race_eth4_h, prevalence = matthews_rate) %>%
  select(race, prevalence) %>%
  mutate(type = "ADRD (Matthews rates)",
         race = recode(race, !!!race_relabel),
         race = factor(race, levels = c("White","Hispanic","Asian","Other")))
prev_data <- bind_rows(prev_race, adrd_race)
ggplot(prev_data, aes(x = race, y = prevalence, fill = race)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.1f%%", prevalence)),
            vjust = -0.5, size = 4.5, fontface = "bold") +
  facet_wrap(~type) +
  scale_fill_manual(values = race_colors) +
  scale_y_continuous(limits = c(0, 20), labels = function(x) paste0(x, "%")) +
  labs(x = NULL, y = "Estimated Prevalence (%)",
       title = "Figure 1. Estimated AD/ADRD prevalence by race/ethnicity",
       caption = "AD overall: 8.9% | ADRD overall: 10.5%") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
        plot.title.position = "plot",
        strip.text = element_text(face = "bold", size = 13),
        axis.text.x = element_text(size = 11),
        axis.text.y = element_text(size = 11),
        axis.title = element_text(size = 12),
        plot.caption = element_text(size = 11, face = "italic", hjust = 0.5),
        panel.grid.major.x = element_blank())
```

```{r disparity-plot, fig.height=5.5, fig.width=8}
pr_data <- pr_pd %>%
  filter(measure == "Prevalence Ratio") %>%
  mutate(race = recode(race, !!!race_relabel),
         race = factor(race, levels = c("Other", "Asian", "Hispanic")),
         outcome_label = case_when(outcome=="mci_h"~"MCI",
                                   outcome=="ad_h"~"AD",
                                   outcome=="adrd_h"~"ADRD",
                                   TRUE~outcome),
         outcome_label = factor(outcome_label, levels = c("MCI", "AD", "ADRD")))
ggplot(pr_data, aes(x = estimate, y = race, color = race)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = lci, xmax = uci), height = 0.2, linewidth = 0.8) +
  geom_point(size = 3.5) +
  facet_wrap(~outcome_label, ncol = 1) +
  scale_color_manual(values = race_colors[c("Other","Asian","Hispanic")]) +
  scale_x_continuous(limits = c(0, 4.2), breaks = seq(0, 4, 1)) +
  labs(x = "Prevalence Ratio (vs. White)", y = NULL,
       title = "Figure 2. AIPSW prevalence ratios by race/ethnicity",
       caption = "95% bootstrap CIs. Reference group: White. AIPSW = Augmented Inverse\nProbability of Selection Weighting. MCI = Mild Cognitive Impairment.") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
        plot.title.position = "plot",
        legend.position = "none",
        strip.text = element_text(face = "bold", size = 13),
        axis.text.y = element_text(size = 11),
        axis.text.x = element_text(size = 11),
        axis.title = element_text(size = 12),
        plot.caption = element_text(size = 10, face = "italic", hjust = 0.5))
```

```{r referral-plot, fig.height=3, fig.width=8}
referral_data <- tibble(
  source = factor(c("ADC clinician/staff","Self-referral","Registry/research",
                     "Outside physician","Other"),
                  levels = rev(c("ADC clinician/staff","Self-referral",
                                  "Registry/research","Outside physician","Other"))),
  pct_sample = c(41, 28, 15, 5, 11),
  pct_impaired = c(69, 38, 30, 69, 48))
ggplot(referral_data, aes(x = pct_impaired, y = source)) +
  geom_col(fill = "#8C1515", width = 0.6) +
  geom_text(aes(label = paste0(pct_impaired, "%")),
            hjust = -0.3, size = 5, fontface = "bold") +
  geom_vline(xintercept = 10, linetype = "dashed", color = "#B1040E", linewidth = 1) +
  annotate("text", x = 10, y = 0.35, label = "Pop. rate (~10%)",
           size = 3.5, color = "#B1040E", hjust = 0.5, fontface = "bold.italic") +
  scale_x_continuous(limits = c(0, 85), labels = function(x) paste0(x, "%")) +
  coord_cartesian(clip = "off") +
  labs(x = "% with Cognitive Impairment", y = NULL,
       title = "Figure 3. Cognitive impairment by referral pathway") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
        plot.title.position = "plot",
        axis.text.y = element_text(size = 11),
        axis.text.x = element_text(size = 11),
        axis.title = element_text(size = 12),
        panel.grid.major.y = element_blank(),
        plot.margin = margin(t = 5, r = 10, b = 15, l = 5))
```

```{r enrichment-plot, fig.height=2.5, fig.width=8}
enrich_data <- tibble(
  group = factor(c("White","Hispanic","Asian","Other"),
                 levels = rev(c("White","Hispanic","Asian","Other"))),
  ratio = c(4.8, 3.8, 4.5, 4.0))
ggplot(enrich_data, aes(x = ratio, y = group, fill = group)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = paste0(ratio, "x")),
            hjust = -0.3, size = 5.5, fontface = "bold") +
  scale_fill_manual(values = race_colors[rev(c("White","Hispanic","Asian","Other"))]) +
  scale_x_continuous(limits = c(0, 7)) +
  labs(x = "Enrichment Ratio (ADRC Prevalence / Population Prevalence)", y = NULL,
       title = "Figure 4. Clinical enrichment ratio by race/ethnicity") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
        plot.title.position = "plot",
        axis.text.y = element_text(size = 11),
        axis.text.x = element_text(size = 11),
        axis.title = element_text(size = 12),
        panel.grid.major.y = element_blank())
```

# Conclusions & Future Directions

- We estimate AD prevalence at **8.9%** among adults 65+ in the three-county catchment, highest among Hispanic (11.6%) and lowest among Asian older adults (5.3%).

- We estimate ADRD prevalence at **10.5%** overall, highest among White (15.9%) and Hispanic (13.8%) older adults.

- Clinical enrichment is pervasive across all referral pathways. Demographic reweighting **does not enable population prevalence estimation** from ADRC data alone.

- However, approximately constant enrichment across groups supports the use of reweighting to estimate **racial and ethnic disparities**.

- **Black adults are underrepresented** in the ADRC (n=12 at 65+). Expanding recruitment is a priority.

### Limitations

- The hybrid model assumes constant enrichment across groups. We tested this empirically (3.8x--4.8x) but with limited precision.
- The Hispanic (n=58) and Other (n=18) ADRC subgroups yield wide confidence intervals.
- The ADRC's etiological composition (~45% Lewy body disease) reflects Stanford's research focus and may not generalize.
- We collapsed Black participants into Other due to small sample size (n=12), precluding Black-White disparity estimation.
- BRFSS excludes institutionalized adults, among whom ADRD prevalence is higher.
- Dhana et al. estimated the CHAP coefficients in a Chicago population; these may not transfer perfectly to the Bay Area.
- The NACC UDS does not capture several covariates used in prior generalizability analyses (income, self-rated health, functional limitations).

### Future Directions

- Expand recruitment of underrepresented groups, particularly Black older adults.
- Disaggregate the Asian category given known heterogeneity across subgroups (Mayeda et al. 2016).
- Supplement NACC data with site-level collection of covariates (e.g., income, functional limitations) not captured in the UDS.
- Replicate this generalizability framework across other ADRCs to assess consistency of enrichment patterns.

### References

1. Dhana K, et al. (2023). Estimated prevalence of Alzheimer's disease by state and county. *Alzheimer's & Dementia*, 19(S10).
2. Matthews KA, et al. (2019). Racial and ethnic estimates of Alzheimer's disease and related dementias in the United States. *Alzheimer's & Dementia*, 15(1), 17--24.
3. Mayeda ER, et al. (2016). Inequalities in dementia incidence between six racial and ethnic groups. *Alzheimer's & Dementia*, 12(3), 216--224.
4. Hayes-Larson E, et al. (2022). Generalizing KHANDLE study findings on racial/ethnic differences in cognitive trajectories. *Alzheimer's & Dementia*, 18(8), 1436--1446.
5. Dahabreh IJ, et al. (2019). Generalizing causal inferences from randomized trials. *Biometrics*, 75(2), 685--694.

### Acknowledgments

Supported by the Stanford Alzheimer's Disease Research Center (NIA P30 AG066515).
