---
title: "Local Burden and Demographic Disparities in Alzheimer's Disease and Related Dementias: A Generalizability Analysis of the Stanford ADRC"
author:
  - name: Tracy Lam-Hine, DrPH, MBA
    affil: 1
  - name: Lisa Goldman-Rosas, PhD, MPH
    affil: "1,2"
  - name: Victor W. Henderson, MD, MS
    affil: "1,3"
affiliation:
  - num: 1
    address: Department of Epidemiology and Population Health, Stanford University School of Medicine
  - num: 2
    address: Department of Primary Care and Population Health, Stanford University School of Medicine
  - num: 3
    address: Department of Neurology and Neurological Sciences, Stanford University School of Medicine
# Poster sizing -- update once guidelines are available
poster_height: "36in"
poster_width: "48in"
# Style
primary_colour: "#8C1515"    # Stanford cardinal
secondary_colour: "#4D4F53"  # Stanford cool grey
accent_colour: "#B1040E"
title_textsize: "60pt"
author_textsize: "36pt"
affiliation_textsize: "24pt"
body_textsize: "28pt"
caption_textsize: "20pt"
column_numbers: 3
output:
  posterdown::posterdown_html:
    self_contained: true
    pandoc_args: --mathjax
bibliography: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.align = "center", out.width = "100%"
)

library(tidyverse)
library(knitr)
library(kableExtra)

# Paths
results_dir <- "/Users/lamhine/Documents/GitHub/gen_sadrc_clin/03_results"

# Load results
hybrid     <- readRDS(file.path(results_dir, "12_hybrid_prevalence.rds"))
pr_pd      <- readRDS(file.path(results_dir, "09_pr_pd_results.rds"))
aipsw      <- readRDS(file.path(results_dir, "07_aipsw_results.rds"))
enrichment <- read.csv(file.path(results_dir, "13_enrichment_diagnostics.csv"))

# Color palette (accessible, Stanford-adjacent)
race_colors <- c(
  "White NH"  = "#8C1515",  # Stanford cardinal
  "Hispanic"  = "#E98300",  # warm orange
  "Asian NH"  = "#007C92",  # teal
  "NH Other"  = "#53565A"   # grey
)
```

```{css, echo=FALSE}
/* Prevent section headers from orphaning at column breaks */
h1, h2, h3 {
  break-after: avoid;
}
/* Enrichment: bullets left, table right — table vertically centered */
.enrichment-wrap {
  display: flex;
  gap: 0.5em;
  align-items: center;
}
.enrichment-wrap .enrichment-bullets {
  flex: 1.2;
  min-width: 0;
}
.enrichment-wrap .enrichment-tbl {
  flex: 0.8;
  min-width: 0;
}
/* Consistent list spacing across all sections */
.poster_body li {
  margin-bottom: 0.5em;
}
```

# Background

- **Alzheimer's disease (AD) and AD-related dementias (ADRD)** affect over 6 million Americans, but local estimates of disease burden -- particularly for diverse communities -- remain limited.

- **National prevalence models often lack estimates for Asian Americans**, a critical gap for the Bay Area where Asian Americans represent a substantial share of the older adult population.

- Alzheimer's Disease Research Centers provide rich clinical data, but enroll a higher proportion of people with cognitive impairment than exists in the general population. This **clinical enrichment** complicates efforts to generalize ADRC findings to the broader community.

```{r dag, fig.height=2.8, fig.width=6, fig.align='center', out.width="75%"}
# Selection DAG: X → Y, X → S ← Y
# S is a collider — conditioning on S=1 (ADRC) distorts the X-Y relationship

dag_nodes <- tibble(
  label = c(
    "X\n(Demographics)",
    "Y\n(Cognitive Status)",
    "S = 1\n(ADRC Participation)"
  ),
  x = c(0, 2, 1),
  y = c(1, 1, 0)
)

dag_edges <- tibble(
  x = c(0, 0, 2),
  y = c(1, 1, 1),
  xend = c(2, 1, 1),
  yend = c(1, 0, 0),
  label = c("Causal", "Selection", "Selection\n(enrichment)")
)

ggplot() +
  # Edges (arrows)
  geom_segment(data = dag_edges,
               aes(x = x, y = y, xend = xend, yend = yend),
               arrow = arrow(length = unit(0.15, "inches"), type = "closed"),
               linewidth = 0.8, color = "#4D4F53") +
  # Edge labels
  geom_text(data = dag_edges,
            aes(x = (x + xend)/2 + c(0, -0.22, 0.22),
                y = (y + yend)/2 + c(0.12, 0, 0),
                label = label),
            size = 3.5, color = "#4D4F53", fontface = "italic") +
  # Nodes
  geom_point(data = dag_nodes, aes(x = x, y = y),
             size = 18, shape = 21, fill = "white", color = "#8C1515",
             stroke = 1.5) +
  # Node labels
  geom_text(data = dag_nodes, aes(x = x, y = y, label = label),
            size = 3.2, fontface = "bold", lineheight = 0.85) +
  # Box around S to indicate conditioning
  annotate("rect", xmin = 0.55, xmax = 1.45, ymin = -0.32, ymax = 0.32,
           fill = NA, color = "#B1040E", linewidth = 1.2, linetype = "dashed") +
  annotate("text", x = 1, y = -0.42, label = "Conditioned on (observed sample)",
           size = 2.8, color = "#B1040E", fontface = "italic") +
  coord_fixed(ratio = 0.7, xlim = c(-0.6, 2.6), ylim = c(-0.55, 1.3)) +
  theme_void()
```

### Research Questions

1. What is the estimated prevalence of AD and ADRD among adults aged 65+ in the Stanford ADRC's three-county catchment area (Alameda, San Mateo, and Santa Clara counties)?

2. After accounting for differences in age, sex, education, and health conditions, are there disparities in cognitive outcomes across racial/ethnic groups?

3. How does the ADRC's clinical enrichment shape its sample, and what does this mean for interpreting ADRC-based findings?

# Data & Methods

```{r data-sources-table}
data_sources <- tibble(
  Source = c("**Stanford ADRC**", "**California BRFSS**",
             "**Dhana et al. 2023**", "**Matthews et al. 2019**"),
  Description = c(
    "Clinical assessments (NACC UDS v3); 529 adults aged 65+; 2015--2024",
    "Population-based telephone survey; 2,150 adults aged 65+ in 3 counties; 2015--2024",
    "Published AD prevalence coefficients from the Chicago Health and Aging Project",
    "Published ADRD prevalence rates by race, age, and sex from Medicare data"
  ),
  Role = c(
    "Clinical outcomes; local race/ethnicity associations",
    "Target population demographics",
    "Base model for AD prevalence estimation",
    "ADRD prevalence via indirect standardization"
  )
)

kable(data_sources, format = "html", escape = FALSE) %>%
  kable_styling(font_size = 26, full_width = TRUE,
                bootstrap_options = c("condensed", "striped")) %>%
  column_spec(1, bold = TRUE, width = "20%") %>%
  column_spec(2, width = "45%") %>%
  column_spec(3, width = "35%")
```

**Question 1 -- Local AD Prevalence (Hybrid Model):** We estimated AD prevalence using the following logistic model applied to individual-level BRFSS data with survey weights:

$$\text{logit}(P(\text{AD}_i)) = \hat{\alpha}^{D} + \hat{\beta}_{\text{age}}^{D} x_{\text{age},i} + \hat{\beta}_{\text{sex}}^{D} x_{\text{sex},i} + \hat{\beta}_{\text{edu}}^{D} x_{\text{edu},i} + \hat{\beta}_{\text{race}}^{A} x_{\text{race},i}$$

where coefficients marked $D$ are from the Dhana et al. (2023) population-based CHAP cohort and $\hat{\beta}_{\text{race}}^{A}$ (the log-odds ratios for AD comparing each racial/ethnic group to White NH) are estimated from the Stanford ADRC. Using case-control study principles, we verified the transportability of these odds ratios by testing for variation in the degree of clinical oversampling across racial/ethnic groups. White NH serves as the reference group for the race coefficients, so no substitution is needed for that group. For ADRD, we applied published Medicare-based prevalence rates by race, age, and sex (Matthews et al. 2019) to the local demographic distribution via indirect standardization.

**Question 2 -- Disparities:** We used augmented inverse probability of selection weighting (AIPSW), a statistical reweighting method that adjusts for demographic and health differences between the ADRC and the local population, to estimate prevalence ratios across racial/ethnic groups (1,000 bootstrap replicates $\times$ 40 multiply imputed datasets).

**Question 3 -- Enrichment:** We characterized the ADRC's referral pathways and compared ADRC-derived odds ratios to published population-based estimates.

# Results

### Local AD/ADRD Prevalence

```{r prevalence-plot, fig.height=5.5, fig.width=8}
prev_race <- hybrid$by_race %>%
  select(race_eth4_h, hybrid_prev) %>%
  rename(race = race_eth4_h, prevalence = hybrid_prev) %>%
  mutate(
    type = "AD (hybrid model)",
    race = factor(race, levels = c("White NH", "Hispanic", "Asian NH", "NH Other"))
  )

adrd_race <- hybrid$adrd_by_race %>%
  rename(race = race_eth4_h, prevalence = matthews_rate) %>%
  select(race, prevalence) %>%
  mutate(
    type = "ADRD (Matthews rates)",
    race = factor(race, levels = c("White NH", "Hispanic", "Asian NH", "NH Other"))
  )

prev_data <- bind_rows(prev_race, adrd_race)

ggplot(prev_data, aes(x = race, y = prevalence, fill = race)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.1f%%", prevalence)),
            vjust = -0.5, size = 5, fontface = "bold") +
  facet_wrap(~type) +
  scale_fill_manual(values = race_colors) +
  scale_y_continuous(limits = c(0, 20), labels = function(x) paste0(x, "%")) +
  labs(
    x = NULL, y = "Estimated Prevalence (%)",
    caption = "AD overall: 8.9% | ADRD overall: 10.5%"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    strip.text = element_text(face = "bold", size = 16),
    axis.text.x = element_text(size = 13, face = "bold"),
    plot.caption = element_text(size = 13, face = "italic", hjust = 0.5),
    panel.grid.major.x = element_blank()
  )
```

Hispanic older adults had the highest estimated AD prevalence (11.6%), while Asian Americans had the lowest (5.3%) -- a finding that national models lacking an Asian category cannot produce.

### Demographic Disparities

```{r disparity-plot, fig.height=6, fig.width=8}
pr_data <- pr_pd %>%
  filter(measure == "Prevalence Ratio") %>%
  mutate(
    race = factor(race, levels = c("NH Other", "Asian NH", "Hispanic")),
    outcome_label = case_when(
      outcome == "mci_h" ~ "MCI",
      outcome == "ad_h" ~ "AD",
      outcome == "adrd_h" ~ "ADRD",
      TRUE ~ outcome
    ),
    outcome_label = factor(outcome_label, levels = c("MCI", "AD", "ADRD"))
  )

ggplot(pr_data, aes(x = estimate, y = race, color = race)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = lci, xmax = uci), height = 0.2, linewidth = 0.8) +
  geom_point(size = 3.5) +
  facet_wrap(~outcome_label, ncol = 1) +
  scale_color_manual(values = race_colors[c("NH Other", "Asian NH", "Hispanic")]) +
  scale_x_continuous(limits = c(0, 4.2), breaks = seq(0, 4, 1)) +
  labs(
    x = "Prevalence Ratio (vs. White NH)",
    y = NULL,
    caption = "AIPSW estimates with 95% bootstrap CIs. Reference: White NH."
  ) +
  theme_minimal(base_size = 16) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 16),
    axis.text.y = element_text(size = 14, face = "bold"),
    plot.caption = element_text(size = 12, face = "italic", hjust = 0.5)
  )
```

Point estimates suggest elevated risk for Hispanic and NH Other groups, but confidence intervals are wide, reflecting modest ADRC subgroup sample sizes.

### Clinical Enrichment

<div class="enrichment-wrap">
<div class="enrichment-bullets">

Approximately **52% of ADRC participants** had cognitive impairment at baseline, compared with an estimated 10% in the general 65+ population (Matthews et al. 2019), representing an approximately 5-fold clinical enrichment. This enrichment was present across all referral pathways, including self-referrals (38% impaired). Enrichment ratios were approximately constant across racial/ethnic groups (White NH: 4.8x, Hispanic: 3.8x, Asian NH: 4.5x), supporting the transportability of ADRC-derived race/ethnicity odds ratios to the target population.

</div>
<div class="enrichment-tbl">

```{r enrichment-table}
enrich_display <- tibble(
  `Referral Source` = c(
    "ADC clinician/staff", "Self-referral",
    "Registry/research", "Outside physician", "Other"
  ),
  `% Sample` = c("41%", "28%", "15%", "5%", "11%"),
  `% Impaired` = c("69%", "38%", "30%", "69%", "48%")
)

kable(enrich_display, format = "html", escape = FALSE) %>%
  kable_styling(font_size = 26, full_width = TRUE,
                bootstrap_options = c("condensed", "striped")) %>%
  column_spec(1, bold = TRUE)
```

</div>
</div>

# Conclusions & Future Directions

- Estimated AD prevalence in the three-county catchment is **8.9%** overall, with the highest burden among Hispanic older adults (11.6%) and the lowest among Asian Americans (5.3%).

- The ADRC's clinical enrichment is substantial and pervasive. ADRC data can provide valuable local race/ethnicity comparisons, but **should not be interpreted as representative of the general population** without statistical adjustment.

- **Black non-Hispanic adults are significantly underrepresented** in the ADRC (n=12 at 65+) relative to the catchment population. Expanding recruitment in this community is a priority.

### Limitations

- The hybrid prevalence model assumes approximately constant clinical enrichment across racial/ethnic groups, which was tested empirically but with limited precision given subgroup sample sizes.
- Hispanic (n=58) and NH Other (n=18) ADRC subgroup estimates carry wide uncertainty.
- The ADRC's high proportion of Lewy body disease (~45% of ADRD) reflects Stanford's research focus.

### Future Directions

- Expand ADRC recruitment of Black and Hispanic older adults to enable adequately powered race-specific analyses.
- Extend the generalizability framework to longitudinal cognitive trajectories and incident dementia.
- Disaggregate the Asian NH category (e.g., Chinese, Filipino, South Asian), given known heterogeneity in dementia risk across Asian subgroups (Mayeda et al. 2016).

### Acknowledgments

This work was supported by the Stanford Alzheimer's Disease Research Center (NIA P30 AG066515).
